Quantitative Evaluation of Touch-Screen Performance with Transition Matrix Measures

### 2015-09-01 flux-entropy-paper/flux-entropy-paper.tex
added todos for the rest of the day
adding lots of plots
updates for the results section

To compare the transition matrices of different interactions we use
two high-level quantities, ``flux'' and ``entropy''. The flux of a
transition matrix $P$ is given by
\begin{equation}
  \mathrm{flux}(P) = \frac{\|P\|_1-\|\mathrm{diag}(P)\|}{\|P\|_1}
\end{equation}
Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
$P$.

Flux measures the ratio of transitions that change
state to self-transitions from a certain state back to itself.
Intuitively, it is a measure of how frequently a group changes state. Flux
returns a value in the interval $[0,1]$ and will return 0 when
participants never change state, and 1 when participants never stay on
the same state for two measurements in a row.

@@

% While musical interactions such as those in
% our corpus are often analysed using qualitative methods, it is
% problematic to

@@

% need to add overall contrasts column to do pairwise t-tests.

@@

@@

@@

>>>>

To compare the transition matrices of different interactions we use
two high-level quantities, ``flux'' and ``entropy''. The flux of a
sequence is defined as the ratio of changed transitions (e.g. state $x$
to state $y$), to self transitions (e.g. state $x$ to state $x$). The
the flux of a transition matrix $P$ can be calculated as:
\begin{equation}
  \mathrm{flux}(P) = 1 - tr(P)
\end{equation}
Where the trace of P, $tr(P)$ is given by the vector 1-norm of the
diagonal entries of $P$. 
% Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
% and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
% $P$.
Intuitively, flux is a measure of how frequently a group changes state. Flux
returns a value in the interval $[0,1]$ and will return 0 when
participants never change state, and 1 when participants never stay on
the same state for two measurements in a row.

@@

@@

% todo need to add overall contrasts column to do pairwise t-tests.

@@

Figures \ref{fig:type-section-flux} and \ref{fig:type-section-entropy}
show that the flux and entropy of compositions are much more stable
with respect to section than for improvisations. The stability of
compositions is to be expected since performers are expected to use
the same gestures in each rehearsal and performance of a composition,
it may be that different kinds of compositions could result in
different structural differences. Both measures have a downward trend
through sections in improvisations with Kruskal-Wallis tests showing a
significant effect of section on flux ($\chi^2(2)=5.8, p=0.05$). This
could be explained by performers entering an exploration stage at the
start of an improvisation, where they change gesture frequently to
experiment with different ideas. At the end of an improvisation,
performers may change gesture less frequently while ``winding up'' the
performance.

The gestural change in improvisation can be further explored by
dividing the sessions by performance context. Figures
\ref{fig:context-section-flux}, and \ref{fig:context-section-entropy} show
this division for improvisations only. In rehearsals and studies the
distribution of flux and entropy over sections has a similar shape,
both measures drop in the ending section of studies and after the
beginning section in rehearsals. In performances however, flux drops
after the beginning while entropy drops substantially in the ending
section. This indicates a more constrained performance style in this
closing section, where performers transition between fewer gestures in
the final section of performances.

% todo add the plots here.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/type-section-flux}
  \caption{The flux of compositions and improvisations divided into
    beginning, middle, and end. Compositions remain stable throughout,
    but improvisations show a downward trend.
    \label{fig:type-section-flux}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/type-section-entropy}
  \caption{The entropy of compositions and improvisations divided by
    beginning, middle, and end. Composition entropy tends to move up
    at the end, while flux moves down.
    \label{fig:type-section-entropy}}
\end{figure}

% todo add the results from the plots.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/context-section-flux}
  \caption{The distribution in performances divided by ``beginnings,
    middles, and ends''.
    \label{fig:context-section-flux}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/context-section-entropy}
  \caption{The distribution of entropy in performances divided into
    sections and by performance context.
    \label{fig:context-section-entropy}}
\end{figure}

@@

% todo add anymore results from these plots

@@

In this paper we have extended the transition matrix methodology
previously presented at CHI with two simple matrix measures, flux and
entropy, both of which have been demonstrated to have significant
discriminatory power with regards to a corpus of 150 collaborative
touch-screen interactions. Not only do these measures differentiate
between different types of sessions and the internal structure of
these sessions, but they have explanatory power that has thrown new
light on the musical interactions under examination. HCI researchers
are increasingly concerned with interactions that happen in-the-wild,
where the interactions of many users may be collected without the
opportunity to survey or interview users. We suggest that analysing
interaction-state transitions using matrices and simple measures can
lead to new insights into HCI interactions outside the lab.

% todo blah de blah
